SU($n$) spin models near the Heisenberg critical point

### 2020-04-02 collective_sun/sun_perturbation.tex
added some todo statements drastically simplified the section on perturbation theory started an introduction

@@

\red{TODO: write an introduction}

@@

@@

In order to induce non-trivial dynamics for initially PS states
$\ket\psi\in\M_0$, which are otherwise stationary eigenstates of the
interaction Hamiltonian $H_0$ in \eqref{eq:H_0}, we will add to the
Hamiltonian an operator $\hat\O$ of the general form
\begin{align}
  \hat\O \equiv \sum_{\p{X,\m w}\in\O} X_{\m w},
  &&
  X_{\m w} \equiv \f1{M!} \sum_{k\in\D_N\p{M}} w_k X_k,
  \label{eq:pert}
\end{align}
where $\hat\O$ is constructed from the set $\O\equiv\set{\p{X,\m w}}$
of $M$-spin operators $X$ together with dimension-$M$ (i.e.~$M$-index)
tensors $\m w$; $k=\p{k_1,k_2,\cdots,k_M}$ is a list of of $M$ spins
that $X_k$ acts on; $w_k$ is a scalar entry of $\m w$; and
\begin{align}
  \D_N\p{M} \equiv
  \set{ k \in \ZZ_N^M: \t{all entries of}~k~\t{are distinct} },
  \label{eq:off_diags}
\end{align}
is the strictly off-diagonal part of $\ZZ_N^M$.  The factor of $1/M!$
is included in \eqref{eq:pert} by convention to average over the $M!$
terms addressing any given set of spins.  Due to the permutational
symmetry enforced by the interaction Hamiltonian $H_0$, we will find
it convenient to decompose $\D_N\p{M}=\pi\p{M}\times\C_N\p{M}$, where
$\pi\p{M}$ is the permutation group of $\ZZ_M$, and
\begin{align}
  \C_N\p{M} \equiv \set{ k \subset \ZZ_N: \abs{k} = M },
  \label{eq:choices}
\end{align}
is the set of all subsets (``choices'') of $M$ elements from $\ZZ_N$.
We can then expand
\begin{align}
  X_{\m w} = \f1{M!} \sum_{\sigma\in\pi\p{M}} \sum_{k\in\C_N\p{M}}
  w_{\sigma\p{k}} X_{\sigma\p{k}}.
\end{align}
To simplify language and notation, we will assume throughout this work
that $\hat\O$ is homogeneous in the sense that $M$ is the same for all
$\p{X,\m w}\in\O$.  A generalization to the case of inhomogeneous
operators $\hat\O$ is straightforward.

Single-body ($M=1$) operators correspond to classical external fields,
e.g.~$s_\z$ for a magnetic field, and two-body ($M=2$) operators
correspond to two-body interactions that deviate from $H_0$, such as
$s_\z\otimes s_\z$ interactions in an XXZ model\red{[TODO:cite]} near
the Heisenberg critical point.  While we would be hard-pressed to find
a physical implementation of multi-body spin interactions with $M>2$,
considering this case will be necessary to construct low-lying
eigenstates of $H_0$, which will in turn enable simulating the
dynamics induced by non-perturbative additions $\hat\O$ to the SU($n$)
Heisenberg model, with operator norms $\norm*{\hat\O}$ comparable to
the spectral gap $\Delta_{\t{gap}}$ of $H_0$.

When the operator norm of $\hat\O$ is sufficiently small, namely
$\norm*{\hat\O}<\Delta_{\t{gap}}/2$, we can treat its effect on the PS
manifold $\M_0$ perturbatively.  The effective Hamiltonian
$H_{\t{eff}}$ induced by $\hat\O$ on the ground-state manifold $\M_0$
through second order in perturbation theory
is\cite{bravyi2011schrieffer, perlin2019effective}
\begin{align}
  H_{\t{eff}} = H_{\t{eff}}^{(1)} + H_{\t{eff}}^{(2)},
  &&
  H_{\t{eff}}^{(1)} = \P_0 \hat\O \P_0,
  &&
  H_{\t{eff}}^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 \hat\O \P_\Delta \hat\O \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of
the interaction Hamiltonian $H_0$ with interaction energy $\Delta$
above that of PS manifold $\M_0$.  In particular, $\P_0$ is a
projector onto $\M_0$ itself, and $\E_0=\M_0$.  The first order
effective Hamiltonian $H_{\t{eff}}^{(1)}$ is straightforward to
simplify using the permutational symmetry of $\M_0$:
\begin{align}
  H_{\t{eff}}^{(1)} \EQPS \sum_{\p{X,\m w}\in\O} \bar w\,\col{X},
  \label{eq:H_eff_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to $\M_0$; the
coefficient $\bar w$ is the average of all $w_k$; and the collective
operator $\col{X}$ is a permutation-averaged sum of all $X_k$, i.e.
\begin{align}
  \bar w \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} w_k,
  &&
  \col{X} \equiv \f1{\abs{\pi\p{M}}} \sum_{\sigma\in\pi\p{M}}
  \sum_{k\in\C_N\p{M}} X_{\sigma\p{k}},
\end{align}
with $\abs{\D_N\p{M}}=\prod_{j=0}^{M-1}\p{N-j}$ and
$\abs{\pi\p{M}}=M!$.  An immediate consequence of \eqref{eq:H_eff_1}
is the fact that perturbing SU(2)-symmetric interactions by two-body
$s_\z\otimes s_\z$ interactions yields the one-axis twisting (OAT)
Hamiltonian\cite{kitagawa1993squeezed, ma2011quantum}
$H_{\t{OAT}}=\chi S_\z^2$ at first order in perturbation theory, with
the OAT strength $\chi$ equal to the mean coefficient of the two-body
$s_\z\otimes s_\z$ terms.

The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ is not as
straightforward to simplify due to the presence of a projector
$\P_\Delta$ onto the manifold $\E_\Delta$ of states with excitation
energy $\Delta$.  For each term $X_{\m w}$ of $\hat\O$, this projector
essentially picks off the part of $X_{\m w}$ that is strictly
off-diagonal with respect to the ground- and excited-state manifolds
$\E_0$ and $\E_\Delta$.  To decompose $X_{\m w}$ into components that
generate states of definite excitation energy when acting on PS states
$\M_0$, we pick an arbitrary state $\ket\psi\in\M_0$ and expand (see
Appendix \ref{sec:eigenstates})
\begin{align}
  H_0 X_{\m w} \ket\psi
  = E_0 X_{\m w} \ket\psi + X_{\hat{\m h}\p{\m w}} \ket\psi,
  \label{eq:diagnosis}
\end{align}
where $\hat{\m h}$ is a linear map on the space of dimension-$M$
tensors, determined entirely by the couplings $h_{pq}$ in $H_0$.  The
image $\hat{\m h}\p{\m w}$ of $\m w$ under $\hat{\m h}$ is then itself
a dimension-$M$ tensor that defines the $M$-body operator
$X_{\hat{\m h}\p{\m w}}$.  To write out $\hat{\m h}\p{\m w}$
explicitly, we first define a matrix
$\m h\equiv\sum_{p,q\in\ZZ_N} h_{pq} \op{p}{q}$ of all couplings in
$H_0$ by enforcing that $h_{pq}=h_{qp}$ and $h_{pp}=0$.  We denote a
contraction of $\m h$ with the $a$-th index of $\m w$ by
$\m h_a\p{\m w}$, such that the $k$-th entry of $\m h_a\p{\m w}$ is
\begin{align}
  \m h_a\p{\m w}_k
  \equiv \sum_{\substack{p\in\ZZ_N\\p\notin k}}
  h_{pk_a} w_{k_1 k_2 \cdots k_{a-1} p k_{a+1} \cdots k_M}
  = \sum_{\substack{p\in\ZZ_N\\p\notin k}} h_{pk_a} w_{k_{a:p}},
\end{align}
where if $k=\p{k_1,k_2,\cdots,k_M}$, then $k_{a:p}$ is the same list
but with the $a$-th entry $k_a$ replaced by $p$.  We similarly define
a tensor $\m h_{\t{P}}\p{\m w}$ with entries
\begin{align}
  \m h_{\t{P}}\p{\m w}_k
  \equiv \sum_{\set{a,b}\in\C_M\p{2}} h_{k_ak_b} w_{k_{a\lra b}},
\end{align}
where $k_{a\lra b}$ is the same as $k$ but with the placement of $k_a$
and $k_b$ swapped.  Finally, we define a tensor $\m h_{\t{D}}\p{\m w}$
with entries
\begin{align}
  \m h_{\t{D}}\p{\m w}_k \equiv h^{\t{D}}_k w_k,
  &&
  h^{\t{D}}_k \equiv \sum_{p\in k} h_p
  - \sum_{\set{a,b}\in\C_M\p{2}} h_{k_ak_b},
  &&
  h_p \equiv \sum_{q\in\ZZ_N} h_{pq}.
\end{align}
These definitions allow us to expand
\begin{align}
  \hat{\m h}\p{\m w}
  \equiv \sum_{a\in\ZZ_M} \m h_a\p{\m w}
  + \m h_{\t{P}}\p{\m w} - \m h_{\t{D}}\p{\m w}.
  \label{eq:multi_body_op}
\end{align}
Viewed an operator on the vector space of dimension-$M$ tensors, the
map $\hat{\m h}$ is real and symmetric under transposition, and is
therefore diagonalizable.  Diagonalizing this operator yields
eigenvalues $\Delta$ and eigenvectors $\m u_\Delta$ for which
$\hat{\m h}\p{\m u_\Delta}=\Delta\m u_\Delta$.  Finding eigenvalues
and eigenvectors of $\hat{\m h}$ nominally requires diagonalizing a
matrix of dimensions $\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}$, but
spatial symmetries of $H_0$ and $X_{\m w}$, such as translational
invariance or isotropy, can drastically reduce the complexity of this
eigenvalue problem.  We discuss these reductions in Appendix
\ref{sec:symmetries}.

Defining $\m w_\Delta$ to be the projection of the tensor $\m w$ onto
the $\Delta$-eigenspace of $\hat{\m h}$, we can expand
\begin{align}
  \m w = \sum_\Delta \m w_\Delta,
  &&
  \hat{\m h}\p{\m w} = \sum_\Delta \Delta \m w_\Delta,
\end{align}
and in turn
\begin{align}
  X_{\m w} = \sum_\Delta X_{\m w_\Delta},
  &&
  X_{\hat{\m h}\p{\m w}} = \sum_\Delta \Delta X_{\m w_\Delta}.
\end{align}
Substituting these expansions into \eqref{eq:diagnosis}, we find
\begin{align}
  H_0 X_{\m w} \ket\psi
  = \sum_\Delta\p{E_0 + \Delta} X_{\m w_\Delta} \ket\psi,
\end{align}
which implies that the action of $X_{\m w_\Delta}$ on a PS state
$\ket\psi\in\M_0$ generates a state with definite interaction energy
$E_0+\Delta$.  It follows that
$\P_\Delta X_{\m w} \P_0 = \P_\Delta X_{\m w_\Delta} \P_0 =
X_{\m{w}_\Delta} \P_0$, so
\begin{align}
  H_{\t{eff}}^{(2)} = -\sum_{\Delta\ne0} \f1\Delta
  \sum_{\p{X,\m v},\p{Y,\m w}\in\O}
  \P_0 X_{\m v_\Delta} Y_{\m w_\Delta} \P_0.
\end{align}
Simplifying product $\P_0 X_{\m v_\Delta} Y_{\m w_\Delta} \P_0$
generally yields $r$-body operators for all positive integers
$r\le2M$.  We provide a general prescription for simplifying such
products in Appendix \ref{sec:operator_product}, and explicitly work
out single-body ($M=1$) and two-body ($M=2$) cases in Appendices
\ref{sec:single_pair_prod} and \ref{sec:two_pair_prod}.  If $X$ is a
single-body operator ($M=1$), then
\begin{align}
  H_{\t{eff}}^{(2)}
  = \sum_{\Delta\ne0} \sum_{\p{X,\m v},\p{Y,\m w}\in\O}
  \f{\cov\p{\m v_\Delta,\m w_\Delta}}{\p{N-1}\Delta}
  \p{\col{X}\,\col{Y} - N \col{XY}},
  \label{eq:H_eff_2_1}
\end{align}
where
\begin{align}
  \cov\p{\m v,\m w} \equiv
  \f1N \sum_{p\in\ZZ_N} \p{v_p-\bar v}\p{w_p-\bar w}
\end{align}
is the covariance between $\m v$ and $\m w$.  In the SU(2) case,
\eqref{eq:H_eff_2_1} implies that an inhomogeneous magnetic field
($X\propto s_\z$) yields an OAT Hamiltonian at second order
perturbation theory.

@@

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \draw[ultra thick] (0,0) -- node[above] {$\M_0$} ++(10em,0);
    \draw[ultra thick] (0,3em) -- node[above] {$\M_1$} ++(10em,0);
    \draw[ultra thick] (0,2*3em) -- node[above] {$\M_2$} ++(10em,0);
    \draw[->] (-.2em,0) to [bend left] (-.2em,3em);
    \node (X) at (.2em,0.5*3em) {$\hat\O_1$};
    \draw[->] (-.4em,0) to [bend left] (-.2em,2*3em);
    \node (XY) at (-2em,3em) {$\hat\O_2$};
    \draw (10em+.2em,-.2em) -- (10em+.4em,-.2em)
    -- (10em+.4em,3em+.2em) -- (10em+.2em,3em+.2em);
    \node (cM1) at (10em+1.5em,0.5*3em) {$\col\M_1$};
    \draw (10em+2.4em+.2em,-.4em) -- (10em+2.4em+.4em,-.4em)
    -- (10em+2.4em+.4em,6em+.4em) -- (10em+2.4em+.2em,6em+.4em);
    \node (cM1) at (10em+2.4em+1.5em,3em) {$\col\M_2$};
  \end{tikzpicture}
  \caption{Schematic of low-lying multi-body manifolds.  Starting with
    the permutationally-state manifold $\M_0$, the single-body
    excitation manifold $\M_1$ consists of states accessible via
    single-body perturbations $\hat\O_1$.  The two-body excitation
    manifold $\hat\O_2$ then consists of states accessible via
    two-body perturbations, but not single-body perturbations.  The
    joint manifold $\col\M_M$ consists of all states accessible by
    operators $r$-body operators $\O_r$ with $r\le M$.}
  \label{fig:manifold_sketch}
\end{figure}

@@

Denoting all assignments of $N$ (identical) spins to $n$ (distinct)
states by $\A_n\p{N}$, a suitable basis for $\M_0$ is the set of PS
states $\set{\ket{m}:m\in\A_n\p{N}}$, where $m=\p{m_1,m_2,\cdots,m_n}$
specifies the occupation number $m_\mu$ of each single-spin state
$\mu$.  For consistency with the expression of matrix elements in
\eqref{eq:matrix_element}, for each $\ket{m}\in\M_0$ we define
$\ket{\1,1,m}\equiv\ket{m}$, with $\1$ the identity operator, and we
define $\1_1\equiv\1$.  We now need to determine the set of $r$-body
operators $\set{O}$ and dimension-$r$ tensors $\set{\m u}$ that define
our basis for each $\M_r$.  As an initial guess for $\set{\m u}$, we
can choose the set $\set{\m u_\Delta}_{\m h,r}$ of dimension-$r$
tensors $\m u_\Delta$ that generate eigenstates of $H_0$, i.e.~for
which $\hat{\m h}\p{\m u_\Delta} = \Delta \m u_\Delta$, with
$\hat{\m h}$ defined in \eqref{eq:multi_body_op}.  The set of $r$-body
operators to choose is less obvious, but can be determined by an
exhaustive search over all $r$-body operators relevant to a particular
scenario.  Specifically, the purpose of projecting operators onto
$\col\M_{r_{\t{max}}}$ is to simulate the dynamics induced by some
operator $\hat\O$ of the form in \eqref{eq:pert}, constructed from a
set $\O=\set{\p{X,\m w}}$ containing multi-spin operators $X$.  In
order to find a basis for the {\it relevant} states in $\M_r$, one can
simply consider the set $\set{O}_{\O,r}$ all $r$-spin operators that
can be constructed from products of the multi-spin operators $X$ in
$\O$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Existence of a spectral gap with long-range power-law
  interactions}
\label{sec:spectral_gap}

@@

size of $\hat{\m h}$ from $N!_M\times N!_M$ to
${N\choose M}\times{N\choose M}$, where
$N!_M=M!\times{N\choose M}=\prod_{j=0}^{M-1}\p{N-j}$ denotes a falling
factorial.  We discuss further reduction of the eigenvalue problem in
\eqref{eq:multi_body_problem} by use of additional symmetries in
Section \ref{sec:symmetries}.

Some special cases: if $M=1$, then $h^{\t{D}}_k=h_k$, so
\begin{align}
  \hat{\m h} \stackrel{M=1}{=} \sum_{k\in\ZZ_N} \ket{k}
  \sp{\sum_{\substack{p\in\ZZ_N\\p\ne k}} h_{pk} \bra{p} - h_k \bra{k}}
  = \m h - \diag\v h,
\end{align}
which is precisely what we found in \eqref{eq:single_body_eig}.  If
$M=2$, then $h_{k\ell}^{\t{D}}=h_k+h_\ell-h_{k\ell}$, so
\begin{align}
  \hat{\m h} \stackrel{M=2}{=}
  \sum_{\p{k,\ell}\in\D_N\p{2}} \ket{k,\ell}
  \sp{\sum_{\substack{p\in\ZZ_N\\p\notin\set{k,\ell}}}
    \p{ h_{pk} \bra{p,\ell} + h_{p\ell} \bra{k,p} }
    + h_{k\ell} \bra{\ell,k}
    - \p{h_k + h_\ell - h_{k\ell}} \bra{k,\ell}}.
\end{align}

>>>>

% todo: solve collective model; Casmir operators?

@@

SU($n$) symmetries play an important role in contemporary physics.
Underpinning much of high energy physics, the celebrated SU($n$) gauge
theory known as Yang-Mills theory is central to our understanding of
the electroweak and strong forces.  Extensions of Yang-Mills and
SU($n$) symmetry feature in the most well-studied examples of
holographic duality\cite{maldacena1999largen} and the connection
between entanglement and gravity\cite{ryu2006holographic} through the
anti-de Sitter / conformal field theory (AdS/CFT) correspondence.  In
a condensed matter setting, SU(2) appears ubiquitously as a symmetry
of the Hubbard model, with important consequences for the study of
quantum magnetism and high temperature
superconductivity\cite{lee2006doping}.  The extension of SU(2) spin
models to the case of SU($n$) has led to predictions of exotic new
phases of matter\cite{hermele2009mott, hermele2011topological,
  chen2016syntheticgaugefield, nataf2016chiral}, as well as the
potential to perform universal topological quantum
computation\cite{freedman2004class, nayak2008nonabelian}.
Furthermore, the introduction of disorder into SU($n$) spin models has
opened avenues for studying quantum chaos and information
scrambling\cite{sachdev1993gapless, bentsen2019integrable}.

Last but not least, SU($n$) symmetries appear naturally in the physics
of atomic, molecular, and optical (AMO)
systems\cite{beverland2016realizing, cazalilla2014ultracold,
  gorshkov2010twoorbital, scazza2014observation, taie2012su,
  zhang2014spectroscopic, goban2018emergence, perlin2019effective},
enabling their experimental study with an exquisite level of control.
These symmetries typically appear through the independence of
inter-atomic interaction parameters on the $n$ nuclear spin states of
the individual atoms involved, with e.g.~$n=10$ in the case of
${}^{87}$Sr.

\red{[finish introduction]}

\vspace{3cm}

@@

% todo: give example of SU(2)

@@

% todo: simplify, give examples at every step
% todo: consider special case before a general one
% todo: new symbol for X_w, e.g. O(X,w)

In order to induce non-trivial dynamics for initially PS states
$\ket\psi\in\M_0$, which are otherwise stationary eigenstates of the
interaction Hamiltonian $H_0$ in \eqref{eq:H_0}, we will add to the
Hamiltonian an operator $\hat\O$ of the general form
\begin{align}
  \hat\O \equiv \sum_{\p{X,\m w}\in\O} X_{\m w},
  &&
  X_{\m w} \equiv \f1{M!} \sum_{k\in\D_N\p{M}} w_k X_k,
  \label{eq:pert}
\end{align}
where $\hat\O$ is constructed from the set $\O\equiv\set{\p{X,\m w}}$
of $M$-spin operators $X$ together with dimension-$M$ (i.e.~$M$-index)
tensors $\m w$; $k=\p{k_1,k_2,\cdots,k_M}$ is a list of of $M$ spins
that $X_k$ acts on; $w_k$ is a scalar entry of $\m w$; and
\begin{align}
  \D_N\p{M} \equiv
  \set{ k \in \ZZ_N^M: \t{all entries of}~k~\t{are distinct} },
  \label{eq:off_diags}
\end{align}
is the strictly off-diagonal part of $\ZZ_N^M$.  The factor of $1/M!$
is included in \eqref{eq:pert} by convention to average over the $M!$
terms addressing any given set of spins.  To simplify language and
notation, we will assume throughout this work that $\hat\O$ is
homogeneous in the sense that $M$ is the same for all
$\p{X,\m w}\in\O$.  A generalization to the case of inhomogeneous
operators $\hat\O$ is straightforward.

Single-body ($M=1$) operators correspond to classical external fields,
e.g.~$s_\z$ for a magnetic field, and two-body ($M=2$) operators
correspond to two-body interactions that deviate from $H_0$, such as
$s_\z\otimes s_\z$ interactions in an XXZ model near the Heisenberg
critical point.  While we would be hard-pressed to find a physical
implementation of multi-body spin interactions with $M>2$, considering
this case will be necessary to construct low-lying eigenstates of
$H_0$, which will in turn enable simulating the dynamics induced by
non-perturbative additions $\hat\O$ to the SU($n$) Heisenberg model,
with operator norms $\norm*{\hat\O}$ comparable to the spectral gap
$\Delta_{\t{gap}}$ of $H_0$.

When the operator norm of $\hat\O$ is sufficiently small, namely
$\norm*{\hat\O}<\Delta_{\t{gap}}/2$, we can treat its effect on the PS
manifold $\M_0$ perturbatively.  The effective Hamiltonian
$H_{\t{eff}}$ induced by $\hat\O$ on the ground-state manifold $\M_0$
through second order in perturbation theory
is\cite{bravyi2011schrieffer, perlin2019effective}
\begin{align}
  H_{\t{eff}} = H_{\t{eff}}^{(1)} + H_{\t{eff}}^{(2)},
  &&
  H_{\t{eff}}^{(1)} = \P_0 \hat\O \P_0,
  &&
  H_{\t{eff}}^{(2)} = - \sum_{\Delta\ne0}
  \f1\Delta \P_0 \hat\O \P_\Delta \hat\O \P_0,
  \label{eq:H_eff}
\end{align}
where $\P_\Delta$ is a projector onto the eigenspace $\E_\Delta$ of
the interaction Hamiltonian $H_0$ with interaction energy $\Delta$
above that of PS manifold $\M_0$.  In particular, $\P_0$ is a
projector onto $\M_0$ itself, and $\E_0=\M_0$. The first order
effective Hamiltonian $H_{\t{eff}}^{(1)}$ simply projects the
perturbation $\hat\O$ onto the permutationally-symmetric manifold
$\M_0$, and takes the form
\begin{align}
  H_{\t{eff}}^{(1)} \EQPS \sum_{\p{X,\m w}\in\O} \bar w\,\col{X},
  \label{eq:H_eff_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to $\M_0$; the
coefficient $\bar w$ is the average of all $w_k$; and the collective
operator $\col{X}$ is a sum of all $X_k$ averaged over permutations of
any particular choice of spins $k\in\D_N\p{M}$.  To write out the
collective operator $\col{X}$ explicitly, we define the set \begin{align}
  \C_N\p{M} \equiv \set{ k \subset \ZZ_N: \abs{k} = M },
  \label{eq:choices}
\end{align}
is the set of all subsets (``choices'') of $M$ elements from $\ZZ_N$,
and the set $\pi\p{M}$ of all permutations of $M$ elements, in terms
of which
\begin{align}
  \bar w \equiv \f1{\abs{\D_N\p{M}}}
  \sum_{k\in\D_N\p{M}} w_k,
  &&
  \col{X} \equiv \f1{\abs{\pi\p{M}}} \sum_{\sigma\in\pi\p{M}}
  \sum_{k\in\C_N\p{M}} X_{\sigma\p{k}},
\end{align}
where $\abs{\D_N\p{M}}=\prod_{j=0}^{M-1}\p{N-j}$ and
$\abs{\pi\p{M}}=M!$.  Here $k\in\C_N\p{M}$ defines which spins
$X_{\sigma\p{k}}$ acts on, while $\sigma\in\pi\p{M}$ determines which
tensor factor of $X$ each of these spins is associated with.  An
immediate consequence of \eqref{eq:H_eff_1} is the fact that
perturbing SU(2)-symmetric interactions by two-body $s_\z\otimes s_\z$
interactions yields the one-axis twisting (OAT)
Hamiltonian\cite{kitagawa1993squeezed, ma2011quantum}
$H_{\t{OAT}}=\chi S_\z^2$ at first order in perturbation theory, with
the OAT strength $\chi$ equal to the mean coefficient of the two-body
$s_\z\otimes s_\z$ terms.

The second order effective Hamiltonian $H_{\t{eff}}^{(2)}$ is not as
straightforward to simplify due to the presence of a projector
$\P_\Delta$ onto the manifold $\E_\Delta$ of states with excitation
energy $\Delta$.  For each term $X_{\m w}$ of $\hat\O$, this projector
essentially picks off the part of $X_{\m w}$ that is strictly
off-diagonal with respect to the ground- and excited-state manifolds
$\E_0$ and $\E_\Delta$.  We therefore need to decompose $X_{\m w}$
into components that generate states of definite excitation energy
when acting on PS states $\ket\psi\in\M_0$.  The SU($n$) symmetry of
$H_0$ enables such a decomposition of the form
\begin{align}
  H_0 X_{\m w}\ket\psi
  = \sum_\Delta \p{E_0+\Delta} X_{\m w_\Delta} \ket\psi,
\end{align}
where, thinking of the tensor $\m w$ as a $\abs{\D_N\p{M}}$-component
vector, the tensor $\m w_\Delta$ can be found found by diagonalizing a
matrix $\hat{\m h}$ of dimensions
$\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}\sim N^M\times N^M$, and
projecting $\m w$ onto the eigenspace of $\hat{\m h}$ with eigenvalue
$\Delta$ (see Appendix \ref{sec:eigenstates}).  The matrix
$\hat{\m h}$ is determined entirely by the couplings couplings
$h_{pq}$ of $H_0$.  While the computational cost of diagonalizing
$\hat{\m h}$ is nominally $O\p{N^{3M}}$, spatial symmetries of $H_0$
and $X_{\m w}$ such as translational invariance or isotropy can
drastically reduce the complexity of this eigenvalue problem.  We
discuss this reduction in Appendix \ref{sec:symmetries}.

Equipped with the decomposition of $X_{\m w}$ into terms that generate
states of definite excitation energy, we can expand
\begin{align}
  H_{\t{eff}}^{(2)} = -\sum_{\Delta\ne0} \f1\Delta
  \sum_{\p{X,\m v},\p{Y,\m w}\in\O}
  \P_0 X_{\m v_\Delta} Y_{\m w_\Delta} \P_0.
\end{align}
Simplifying product $\P_0 X_{\m v_\Delta} Y_{\m w_\Delta} \P_0$
generally yields $r$-body operators for all positive integers
$r\le2M$.  We provide a general prescription for simplifying such
products in Appendix \ref{sec:operator_product}, and explicitly work
out single-body ($M=1$) and two-body ($M=2$) cases in Appendices
\ref{sec:single_pair_prod} and \ref{sec:two_pair_prod}.  If $X$ is a
single-body operator ($M=1$), then
\begin{align}
  H_{\t{eff}}^{(2)}
  = \sum_{\Delta\ne0} \sum_{\p{X,\m v},\p{Y,\m w}\in\O}
  \f{\cov\p{\m v_\Delta,\m w_\Delta}}{\p{N-1}\Delta}
  \p{\col{X}\,\col{Y} - N \col{XY}},
  \label{eq:H_eff_2_1}
\end{align}
where
\begin{align}
  \cov\p{\m v,\m w} \equiv
  \f1N \sum_{p\in\ZZ_N} \p{v_p-\bar v}\p{w_p-\bar w}
\end{align}
is the covariance between $\m v$ and $\m w$.  In the SU(2) case,
\eqref{eq:H_eff_2_1} implies that an inhomogeneous magnetic field
($X\propto s_\z$) yields an OAT Hamiltonian at second order
perturbation theory.

@@

% give examples of these manifolds in SU(2)

@@

\begin{figure}
  \centering
  \begin{tikzpicture}
    \draw[ultra thick] (0,0) -- node[above] {$\M_0$} ++(10em,0);
    \draw[ultra thick] (0,3em) -- node[above] {$\M_1$} ++(10em,0);
    \draw[ultra thick] (0,2*3em) -- node[above] {$\M_2$} ++(10em,0);
    \draw[->] (-.2em,0) to [bend left] (-.2em,3em);
    \node (X) at (.2em,0.5*3em) {$\hat\O_1$};
    \draw[->] (-.4em,0) to [bend left] (-.2em,2*3em);
    \node (XY) at (-2em,3em) {$\hat\O_2$};
    \draw (10em+.2em,-.2em) -- (10em+.4em,-.2em)
    -- (10em+.4em,3em+.2em) -- (10em+.2em,3em+.2em);
    \node (cM1) at (10em+1.5em,0.5*3em) {$\col\M_1$};
    \draw (10em+2.4em+.2em,-.4em) -- (10em+2.4em+.4em,-.4em)
    -- (10em+2.4em+.4em,6em+.4em) -- (10em+2.4em+.2em,6em+.4em);
    \node (cM1) at (10em+2.4em+1.5em,3em) {$\col\M_2$};
  \end{tikzpicture}
  \caption{Schematic of low-lying multi-body excitation manifolds.
    Starting with the permutationally-state manifold $\M_0$, the
    single-body excitation manifold $\M_1$ consists of states
    accessible via single-body perturbations $\hat\O_1$.  The two-body
    excitation manifold $\hat\O_2$ then consists of states accessible
    via two-body perturbations, but not single-body perturbations.
    The joint manifold $\col\M_M$ consists of all states accessible by
    operators $r$-body operators $\O_r$ with $r\le M$.}
  \label{fig:manifold_sketch}
\end{figure}

@@

Denoting all assignments of $N$ (identical) spins to $n$ (distinct)
states by $\A_n\p{N}$, a suitable basis for $\M_0$ is the set of PS
states $\set{\ket{m}:m\in\A_n\p{N}}$, where $m=\p{m_1,m_2,\cdots,m_n}$
specifies the occupation number $m_\mu$ of each single-spin state
$\mu$.  For consistency with the expression of matrix elements in
\eqref{eq:matrix_element}, for each $\ket{m}\in\M_0$ we define
$\ket{\1,1,m}\equiv\ket{m}$, with $\1$ the identity operator, and we
define $\1_1\equiv\1$.  We now need to determine the set of $r$-body
operators $\set{O}$ and dimension-$r$ tensors $\set{\m u}$ that define
our basis for each $\M_r$.  As an initial guess for $\set{\m u}$, we
can choose the set $\set{\m u_\Delta}_{\m h,r}$ of dimension-$r$
tensors $\m u_\Delta$ that generate eigenstates of $H_0$.  The set of
$r$-body operators to choose is less obvious, but can be determined by
an exhaustive search over all $r$-body operators relevant to a
particular scenario.  Specifically, the purpose of projecting
operators onto $\col\M_{r_{\t{max}}}$ is to simulate the dynamics
induced by some operator $\hat\O$ of the form in \eqref{eq:pert},
constructed from a set $\O=\set{\p{X,\m w}}$ containing multi-spin
operators $X$.  In order to find a basis for the {\it relevant} states
in $\M_r$, one can simply consider the set $\set{O}_{\O,r}$ all
$r$-spin operators that can be constructed from products of the
multi-spin operators $X$ in $\O$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Existence of a spectral gap with long-range power-law
  interactions}
\label{sec:spectral_gap}
% todo: make a sub subsection of B.1

@@

size of $\hat{\m h}$ from $\abs{\D_N\p{M}}\times\abs{\D_N\p{M}}$ to
$\abs{\C_N\p{M}}\times\abs{\C_N\p{M}}$, where
$\abs{\D_N\p{M}}=M!\times{N\choose M}=\prod_{j=0}^{M-1}\p{N-j}$ and
$\abs{\C_N\p{M}}={N\choose M}$.  We discuss further reduction of the
eigenvalue problem in \eqref{eq:multi_body_problem} by use of
additional symmetries in Section \ref{sec:symmetries}.

Some special cases of practical interest: if $M=1$, then
$h^{\t{D}}_k=h_k$, so
\begin{align}
  \hat{\m h} \stackrel{M=1}{=} \sum_{k\in\ZZ_N} \ket{k}
  \sp{\sum_{\substack{p\in\ZZ_N\\p\ne k}} h_{pk} \bra{p} - h_k \bra{k}}
  = \m h - \diag\v h,
\end{align}
which is precisely what we found in \eqref{eq:single_body_eig}.  If
$M=2$, then $h_{k\ell}^{\t{D}}=h_k+h_\ell-h_{k\ell}$, so
\begin{align}
  \hat{\m h} \stackrel{M=2}{=}
  \sum_{\p{k,\ell}\in\D_N\p{2}} \ket{k,\ell}
  \sp{\sum_{\substack{p\in\ZZ_N\\p\notin\set{k,\ell}}}
    \p{ h_{pk} \bra{p,\ell} + h_{p\ell} \bra{k,p} }
    + h_{k\ell} \bra{\ell,k}
    - \p{h_k + h_\ell - h_{k\ell}} \bra{k,\ell}}.
\end{align}
